Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms that correspond to the first stage and second stage
computations in the original term. The idea is that evaluating the
first monostaged term in a first pass, and then evaluating the second
monostaged term on the results of the first pass, yields the same
result as fully evaluating the original \lang\ program (performing
both first stage and second stage evaluation).  While world
\bbtwo\ terms fully evaluate to values in stage two, complexity arises
because evaluation of \bbonem\ terms may also generate results in the
first stage. (For example, a \bbonem\ term of type $\curr \rmint$
results in an \rmint\ in the first stage.) Handling this asymmetry
between the results of evaluating \bbonem\ terms (which may produce
results at both the first and second stage) and \bbtwo\ terms (which
only produce second stage results) is fundamental to the structure of
the splitting algorithm presented in this section.  By construction,
\bbonep\ terms are monostage and do not require splitting.

>>>>

Splitting statically transforms terms in \lang\ into a pair of
monostaged terms ($f_1$ and $f_2(-)$) that correspond to the first
stage and second stage computations in the original term. Since full
evaluation of world \bbtwo\ terms (both first stage and second stage
evaluation) only produces values in stage two, the idea of splitting
is to generate terms where evaluating $f_1$ in a first pass, and then
evaluating $f_2$ on the result of the first pass, yields the same
value as full evaluation of the original \lang\ program.

Similarly, when splitting \bbonem\ terms, first pass evaluation of
$f_1$ should produce the same value as first stage evaluation of the
original \lang\ term.  However, complexity arises because
\bbonem\ terms may also contain computations to be evaluated in the
second stage. (For example, a \bbonem\ term of type $\fut \rmint$
results in an \rmint\ value in the second stage.) Thus we also want
the second pass evaluation of $f_2$ on the outputs of $f_1$ to yield
the same value as fully evaluating all second stage computations in
the original \bbonem\ term. Finally, when splitting \bbonep\ terms,
which only produce values in the first stage, we only need to compare
the result of $f_1$ with that of the \lang\ term.

Handling the asymmetry between the results of evaluating
\bbonem\ terms (which may produce results at both the first and second
stage) and \bbtwo\ terms (which only produce second stage results) is
fundamental to the structure of the splitting algorithm presented in
this section.
